# The `validate` package

Here is an introductory example of using the `validate` package to validate a data set. We will build on this in future chapters.

## Essential background reading

Ideally we recommend you read [Data Validation Infrastructure for R](https://www.jstatsoft.org/article/view/v097i10) before proceeding. This paper introduces the `validate` package and covers the essential concepts for the rest of this training.

## An introductory example 🐕

First we load some packages.

```{r message=FALSE}
# 1. Load the packages ########################################################
library(readxl)
library(validate)
library(yaml)
library(dplyr)
library(gt)
library(jsonlite)
```

We then read in the data and take a look.

```{r message=FALSE}
# 2. Load the data, validation rules, and code list ###########################
## 2.1 Load the data ----------------------------------------------------------
dog_walks <- readxl::read_excel(
  path = "data/dog_walks.xlsx",
  sheet = "dog_walks")

gt(dog_walks)
```

We have a data set of dog walks; including each dog's name and how many squirrels they spotted during their walk.

The rules for this data set are stored in a CSV file. We read these into R using the `validate::validator` function. This creates a validator object.

```{r message=FALSE}
## 2.2 Load the validation rules -----------------------------------------------
rules <- validate::validator(.data = read.csv(file = "data/dog_walks_rules.csv"))

## 2.3 View the validation rules -----------------------------------------------
gt(as.data.frame(rules))

```

You may notice that some rules refer to `code_list`. These are the valid options for certain columns in the data set. These are defined in a YAML file.

```{r message=FALSE}

## 2.4 Load the code list ------------------------------------------------------
code_list <- yaml::yaml.load_file(input = "data/dog_walks_code_list.yml")

## 2.5 View the code list ------------------------------------------------------
jsonlite::toJSON(code_list, pretty = TRUE)

```

Now we can validate the data by creating a `validation` object.

If you find it helpful, you can think of the `validator` object as an investigator and the `validation` object as the investigator’s report on the dataset it’s validating.

The `summary` function creates a summary of the validation with one row per rule.

```{r message=FALSE}

# 3. Confront the data with the validation rules ##############################
validation <- validate::confront(
  dat = dog_walks
  , x = rules
  , key = "walk_id")

## 3.1 View a summary of the validation ---------------------------------------
validation %>% 
  summary() %>% 
  gt()

```

We can also view validation results in detail, with one row per check. This produces a very large table. So we filter it to show only where rules have failed or could not be verified.

```{r message=FALSE}

## 3.2 View the results of the validation -------------------------------------
validation %>% 
  as.data.frame() %>% 
  filter(value == FALSE | is.na(value)) %>%
  arrange(value) %>%
  gt()
```

Validation is now complete, but at this point only R users can see the results.

The next step is to turn these results into clear, practical feedback to help data providers resolve any issues. We do this by creating two outputs:

-   A validation log
-   Validation cell highlighting
